---
title: "Week 2 Assignments: Logarithms in R"
subtitle: "IBM 6540 — Applied Machine Learning for Marketing"
author: "Jae Jung"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-expand: 1
    toc-location: right-body
    toc-title: "Assignments"
    number-sections: true
    code-fold: show
    code-tools: true
    theme: cosmo
    highlight-style: github
    df-print: paged
    link-external-newwindow: true
    embed-resources: true
execute:
  warning: false
  message: false
  freeze: true
---

# Overview {.unnumbered}

These assignments practice the logarithm operations covered in the Week 2
notebook (Section 2.1 — *Explaining how Income distribution changes from
normal to logged data*).

You will practice:

- Converting dollar values to log scale (`log()`, `log10()`)
- Converting log-scale values back to dollars (`exp()`)
- Computing the geometric standard deviation (GSD)
- Working with the real `mail_data` dataset
- Understanding why the choice of log base doesn't affect modeling

> **Reference:** Week 2 notebook, Section 2.1 —
> <https://jaejungca.github.io/recipes/w02_ds_preprocessing.html>

---

# Setup

Run this chunk first — it loads the packages and recreates the direct-mail
dataset used throughout the course.

```{r setup}
library(tidyverse)

# ── Recreate the direct-mail dataset ─────────────────────────────────────────
set.seed(123)
n <- 3000

mail_data <- tibble(
  customer_id = paste0("C", str_pad(1:n, 5, pad = "0")),
  age = round(rnorm(n, mean = 45, sd = 12)),
  income = round(rlnorm(n, meanlog = 10.8, sdlog = 0.6)),
  recency_days = round(rexp(n, rate = 1 / 60)),
  freq_12mo = rpois(n, lambda = 3),
  avg_order_amt = round(rlnorm(n, meanlog = 4.2, sdlog = 0.5), 2),
  channel = sample(
    c("email", "direct_mail", "digital"),
    n,
    replace = TRUE,
    prob = c(0.5, 0.3, 0.2)
  ),
  region = sample(
    c("West", "South", "Midwest", "Northeast"),
    n,
    replace = TRUE
  ),
  loyalty_tier = sample(
    c("Bronze", "Silver", "Gold", "Platinum"),
    n,
    replace = TRUE,
    prob = c(0.4, 0.3, 0.2, 0.1)
  ),
  responded = rbinom(
    n,
    1,
    prob = plogis(
      -3 +
        0.02 * (age - 45) +
        0.3 * log(income / 50000) +
        0.1 * freq_12mo -
        0.005 * recency_days
    )
  )
) |>
  mutate(
    income = if_else(runif(n) < 0.08, NA_real_, income),
    avg_order_amt = if_else(runif(n) < 0.05, NA_real_, avg_order_amt),
    responded = factor(responded, levels = c(1, 0), labels = c("yes", "no"))
  )

glimpse(mail_data)
```

---

# Assignment L.1 — Dollar to Log

Customer `C00005` has an income of \$233,070.

1. Convert this value to the **natural log**: `log(233070)`
2. Convert this value to **log base 10**: `log10(233070)`
3. Verify both results using R
4. In your own words, why does the recipe use `base = 10` rather than the
   natural log?

::: {.callout-tip title="Hint"}
In R, `log(x)` computes the natural log ($\ln x$) and `log10(x)` computes
log base 10. You can also write `log(x, base = 10)` — this is exactly
what `step_log(..., base = 10)` uses internally.
:::

```{r assignment-L1, eval=FALSE}
# Your code here

```

---

# Assignment L.2 — Log Back to Dollar (Median and Mean)

The income variable was generated with `meanlog = 10.8` and `sdlog = 0.6`.
These parameters describe the distribution of **log(income)**, not income itself.

1. Convert the log-scale **median** (10.8) back to dollars using `exp()`
2. Convert the log-scale **mean** (10.98) back to dollars using `exp()`
3. Verify these match the formulas from the Week 2 notebook:

$$\text{Median} = e^{\mu} = e^{10.8}$$

$$\text{Mean} = e^{\mu + \sigma^2/2} = e^{10.8 + 0.6^2/2}$$

4. Why is the mean higher than the median on the dollar scale?

::: {.callout-tip title="Hint"}
`exp(x)` is the inverse of `log(x)`. If `log(income) = 10.8`, then
`exp(10.8)` gives you income back in dollars. For the mean, compute
`10.8 + 0.6^2 / 2` first, then apply `exp()`.
:::

```{r assignment-L2, eval=FALSE}
# Your code here

```

---

# Assignment L.3 — Geometric Standard Deviation

The **geometric standard deviation (GSD)** describes the typical spread of
income around the median on the original dollar scale.

$$\text{GSD} = e^{\sigma} = e^{0.6}$$

1. Compute the GSD using `exp(0.6)`
2. Using the median income from L.2 (~\$49,021), compute the typical range:
   - Lower bound: `median_income / GSD`
   - Upper bound: `median_income * GSD`
3. Verify this matches the shaded green band in the Week 2 plot
   (~\$26,935 to ~\$89,218)
4. Interpret this range in plain English: what does it tell a marketer about
   the "typical" customer in this dataset?

::: {.callout-tip title="Hint"}
Unlike the regular standard deviation (which adds and subtracts), the GSD
**multiplies and divides** because we are on a multiplicative (log) scale.
:::

```{r assignment-L3, eval=FALSE}
# Your code here

```

---

# Assignment L.4 — Confirm with Real Data

Now verify the theoretical values from L.2 and L.3 against the actual
`mail_data` dataset.

1. Compute the **median** and **mean** of `income`, removing `NA`s with
   `na.rm = TRUE`. How close are they to the theoretical values?
2. Create a new column called `log_income` using `mutate()` and `log()`.
   What are the median and mean of `log_income`?
3. Apply `exp()` to the median and mean of `log_income`. Do you recover
   the original dollar-scale values? Explain why or why not.

::: {.callout-tip title="Hint"}
Use `summarise()` with `median()` and `mean()` to compute summary statistics.
The theoretical and empirical values may differ slightly because
`mail_data` is a finite random sample — not the infinite population the
parameters describe.
:::

```{r assignment-L4, eval=FALSE}
# Step 1: median and mean of income (removing NAs)

# Step 2: create log_income and summarise

# Step 3: back-transform with exp()

```

---

# Assignment L.5 — Base 10 vs. Natural Log

Your recipe uses `step_log(..., base = 10)`. Does the choice of base actually
matter for modeling?

1. For income values of \$10,000, \$100,000, and \$1,000,000, compute both
   `log()` (natural) and `log10()` for each value
2. Present the results as a tibble with columns: `income`, `log_natural`,
   `log_base10`
3. Divide `log_natural` by `log_base10` for each row. What constant do you get?
   (Hint: this constant is `log(10)`)
4. Since the two columns differ only by a constant, does it matter which base
   we use for preprocessing? Explain why or why not in terms of what
   `step_normalize()` does afterward.

::: {.callout-tip title="Hint"}
`log(10)` ≈ 2.303. Multiplying or dividing by a constant shifts and scales
values — but `step_normalize()` centers and scales anyway, so any constant
difference between bases gets absorbed.
:::

```{r assignment-L5, eval=FALSE}
# Step 1 & 2: build the tibble
tibble(
  income = c(10000, 100000, 1000000)
) |>
  mutate(
    log_natural = log(income),
    log_base10 = log10(income)
  )

# Step 3: compute the ratio

# Step 4: your written explanation here (as a comment)
```

---

# Summary

| Assignment | Key operation | R function |
|---|---|---|
| L.1 | Dollar → log | `log()`, `log10()` |
| L.2 | Log → dollar (median & mean) | `exp()` |
| L.3 | Geometric standard deviation | `exp(sdlog)` |
| L.4 | Empirical vs. theoretical values | `mutate()`, `summarise()` |
| L.5 | Base 10 vs. natural log | `log()` / `log10()` ratio |
